# Dimensionality reduction

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

```{r, echo=FALSE, results="hide"}
library(BiocStyle)
library(HDF5Array)
library(SummarizedExperiment)
se.out <- readRDS("objects/qc_mat.rds")
dec <- read.table("objects/hvg_output.txt", stringsAsFactors=FALSE, header=TRUE)
```

Taking the HVGs to define the biological subspace of interest.

```{r}
sig <- dec$FDR <= 0.05
summary(sig)
```

Extracting the HVGs, mean-centering and scaling them.

```{r}
exprs.mat <- assay(se.out, "exprs", withDimnames=FALSE)
exprs.mat <- exprs.mat[match(dec$Ensembl[sig], rowData(se.out)$Ensembl), ]
exprs.mat <- (exprs.mat - dec$mean[sig])/sqrt(dec$total[sig])
```

Randomly selecting 10000 cells and using them for PCA.
We've already centered and scaled, so no need to do that again.

```{r}
set.seed(123456)
options(DelayedArray.block.size=2.5e8)
chosen <- sort(sample(ncol(exprs.mat), 10000)) 
selected <- t(as.matrix(realize(exprs.mat[,chosen])))
pc.out <- prcomp(selected, scale=FALSE, center=FALSE, rank=3)
```

Making a plot of the first two PCs.

```{r pcaplot, fig.width=10, fig.height=10}
par(mfrow=c(1,2))
smoothScatter(pc.out$x[,1], pc.out$x[,2], xlab="PC1", ylab="PC2")
smoothScatter(pc.out$x[,1], pc.out$x[,3], xlab="PC1", ylab="PC3")
```

Computing coordinates for all cells.

```{r}
pc.mat <- as.matrix(t(pc.out$rotation) %*% exprs.mat)
dim(pc.mat)
```

Also getting the percentage of variance explained.

```{r}
pc.var <- pc.out$sdev^2
var.exp <- pc.var[seq_len(nrow(pc.mat))]/sum(pc.var)
saveRDS(list(coords=pc.mat, var=var.exp), file="objects/pc_out.rds")
```

