# Dimensionality reduction

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

```{r, echo=FALSE, results="hide"}
library(BiocStyle)
library(SingleCellExperiment)
sce <- readRDS("objects/sce.rds")
dec <- read.table("objects/hvg_output.txt", stringsAsFactors=FALSE, header=TRUE)
```

Taking the HVGs to define the biological subspace of interest.

```{r}
sig <- dec$FDR <= 0.05 & !is.na(dec$FDR)
summary(sig)
```

Extracting the HVGs and mean-centering them.

```{r}
exprs.mat <- logcounts(sce)
exprs.mat <- exprs.mat[match(dec$Ensembl[sig], rowData(sce)$Ensembl), ]
exprs.mat <- exprs.mat - dec$mean[sig]
```

Performing random PCA on the resulting expression matrix to retain the first 50 PCs.
Here we use the `r Githubpkg("Bioconductor/BigDataAlgorithms")` package that contains a randomized SVD implementation for `DelayedArray` objects.

```{r}
library(BigDataAlgorithms)
set.seed(1000)
options(DelayedArray.block.size=200e6)
svd.out <- rsvd(t(exprs.mat), k=50)
pc.out <- sweep(svd.out$u, 2, svd.out$d, "*")
```

Having a look at the variance explained by the top 50 PCs.

```{r screeplot}
var.exp <- svd.out$d^2/(ncol(exprs.mat)-1)
total.var <- sum(dec$total[sig])
plot(var.exp/total.var*100, xlab="PC number", ylab="% variance explained")
```

Making a plot of the first three PCs.

```{r pcaplot, fig.width=12, fig.height=6}
par(mfrow=c(1,2))
smoothScatter(pc.out[,1], pc.out[,2], xlab="PC1", ylab="PC2")
smoothScatter(pc.out[,1], pc.out[,3], xlab="PC1", ylab="PC3")
```

Storing the PCs in the `SingleCellExperiment` object. 
Also saving the SVD values for future use.

```{r}
attr(pc.out, "percentVar") <- var.exp/total.var
reducedDim(sce, "rPCA") <- pc.out
saveRDS(svd.out, file="objects/rsvd.rds")
```

<!--
Saving the updated SCE object.

```{r}
saveRDS(sce, file="objects/sce.rds")
```
-->
