---
title: "Clustering 1.3M cells with FlowSOM"
author: ""
output: html_document
---

```{r}
suppressPackageStartupMessages(library(TENxBrainData))
suppressPackageStartupMessages(library(scater))
suppressPackageStartupMessages(library(flowCore))
suppressPackageStartupMessages(library(FlowSOM))
suppressPackageStartupMessages(library(scran))
suppressPackageStartupMessages(library(pheatmap))
suppressPackageStartupMessages(library(tibble))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(reshape2))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(Rtsne))
suppressPackageStartupMessages(library(mclust))
```

## Read data

```{r}
tenx <- TENxBrainData()
sce <- readRDS("objects/sce.rds")
tenx <- tenx[, colnames(sce)] # drop 19,672 cells from the raw TENxBrainData
counts(sce) <- counts(tenx) # overwrite inbuilt absolute path
system.time(sce <- scater::normalize(sce)) # generate normalized expression values
```

## Read rSVD results

```{r}
svd <- readRDS("objects/rsvd.rds")
pca <- sweep(svd$u, 2, svd$d, "*")
colnames(pca) <- paste0("PC", seq_len(ncol(pca)))
dim(pca)
```

## Apply FlowSOM to cluster the cells based on the first 50 PCs

```{r}
set.seed(123)
system.time(ff <- flowFrame(exprs = pca))
system.time(fSOM <- FlowSOM::ReadInput(ff, compensate = FALSE, transform = FALSE, 
                                       scale = FALSE, silent = TRUE))
system.time(fSOM <- FlowSOM::BuildSOM(fSOM, silent = TRUE, xdim = 13, ydim = 13))
system.time(metaClustering <- metaClustering_consensus(fSOM$map$codes, k = 16))
```

## Add cluster info to data object

```{r}
colData(sce)$som100 <- fSOM$map$mapping[, 1]
colData(sce)$sommeta <- metaClustering[fSOM$map$mapping[, 1]]
```

## Add rownames

```{r}
rownames(sce) <- paste0(rowData(sce)$Ensembl, ".", rowData(sce)$Symbol)
```


## Plot PCA representation

```{r pca, fig.width = 8, fig.height = 8}
dim(sce)
dim(pca)
cols <- c("#DC050C", "#E8601C", "#7BAFDE", "#1965B0", "#B17BA6",
          "#882E72", "#F1932D", "#F6C141", "#F7EE55", "#4EB265",
          "#90C987", "#CAEDAB", "#777777", "black", "cyan", "pink")
names(cols) <- as.character(seq_len(length(cols)))
dfpca <- data.frame(pca, som100 = factor(colData(sce)$som100), 
                    sommeta = factor(colData(sce)$sommeta), 
                    library_id = factor(colData(sce)$Library),
                    mouse = factor(colData(sce)$Mouse),
                    stringsAsFactors = FALSE)
print(ggplot(dfpca, aes(x = PC1, y = PC2, color = sommeta)) + 
        geom_point(size = 0.5) + scale_color_manual(values = cols) + 
        theme_bw())
print(ggplot(dfpca, aes(x = PC3, y = PC4, color = sommeta)) + 
        geom_point(size = 0.5) + scale_color_manual(values = cols) + 
        theme_bw())
print(ggplot(dfpca, aes(x = PC5, y = PC6, color = sommeta)) + 
        geom_point(size = 0.5) + scale_color_manual(values = cols) + 
        theme_bw())
print(ggplot(dfpca, aes(x = PC7, y = PC8, color = sommeta)) + 
        geom_point(size = 0.5) + scale_color_manual(values = cols) + 
        theme_bw())
print(ggplot(dfpca, aes(x = PC9, y = PC10, color = sommeta)) + 
        geom_point(size = 0.5) + scale_color_manual(values = cols) + 
        theme_bw())
```

Also plot PCA colored by library ID/mouse

```{r pca2}
print(ggplot(dfpca, aes(x = PC1, y = PC2, color = library_id)) + 
        geom_point(size = 0.5) + theme_bw())
print(ggplot(dfpca, aes(x = PC1, y = PC2, color = mouse)) + 
        geom_point(size = 0.5) + theme_bw())
```

## Find marker genes

```{r scran_markers}
system.time(scran_markers_all <- scran::findMarkers(sce, 
                                                    clusters = colData(sce)$sommeta,
                                                    block = NULL, 
                                                    design = NULL,
                                                    direction = "up",
                                                    pval.type = "any",
                                                    assay.type = "logcounts",
                                                    get.spikes = FALSE, 
                                                    log.p = TRUE, 
                                                    lfc = 0.5))
for (i in seq_len(length(scran_markers_all))) {
  print(head(as.data.frame(scran_markers_all[[i]])))
}
scran_markers <- unique(unlist(lapply(scran_markers_all, function(w) {
  rownames(subset(w, Top <= 1))
})))

scescranmarker <- sce[which(rownames(sce) %in% scran_markers), ]
logcounts_scranmarkers <- logcounts(scescranmarker)
rownames(logcounts_scranmarkers) <- rowData(scescranmarker)$Symbol
dfscran <- as.data.frame(logcounts_scranmarkers) %>% tibble::rownames_to_column("gene") %>%
  reshape2::melt() %>% dplyr::left_join(as.data.frame(colData(scescranmarker)) %>% 
                                          tibble::rownames_to_column("variable") %>%
                                          dplyr::select(variable, som100, sommeta)) %>%
  dplyr::mutate(sommeta = factor(sommeta)) %>%
  dplyr::mutate(som100 = factor(som100))
```

Heatmap of inferred marker genes

```{r scran_markers_heatmap, fig.height = 12, fig.width = 10}
dfsumscran <- dfscran %>% dplyr::group_by(gene, som100) %>%
  dplyr::summarize(value = quantile(value, probs = 0.75)) %>%
  tidyr::spread(som100, value) %>% as.data.frame()
rownames(dfsumscran) <- dfsumscran$gene
dfsumscran$gene <- NULL
dfsumscran <- dfsumscran[apply(dfsumscran, 1, sd) > 0, ]
pheatmap::pheatmap(dfsumscran, scale = "row", cluster_rows = TRUE, cluster_cols = TRUE, 
                   annotation_col = data.frame(metaClust = factor(metaClustering), row.names = as.character(seq_len(length(metaClustering)))), 
                   show_colnames = FALSE, show_rownames = TRUE, fontsize_row = 6, 
                   annotation_colors = list(metaClust = cols))

```


## Look at known marker genes. 
Marker genes for cell types in mouse brain were obtained from two recent publications:

- [http://science.sciencemag.org/content/347/6226/1138.full](http://science.sciencemag.org/content/347/6226/1138.full)
- [https://www.nature.com/articles/nn.4216](https://www.nature.com/articles/nn.4216)

```{r markers, fig.width = 24, fig.height = 24}
marker_genes <- read.csv("resources/marker_genes.csv", header = TRUE, as.is = TRUE) %>%
  dplyr::arrange(population) %>% dplyr::select(gene, population) %>%
  dplyr::distinct()

keep <- which(rowData(sce)$Symbol %in% unique(marker_genes$gene))
scemarker <- sce[keep, ]
logcounts_markers <- assays(scemarker)[["logcounts"]]
rownames(logcounts_markers) <- rowData(scemarker)$Symbol
df <- as.data.frame(logcounts_markers) %>% tibble::rownames_to_column("gene") %>%
  reshape2::melt() %>% dplyr::left_join(as.data.frame(colData(scemarker)) %>% 
                                          tibble::rownames_to_column("variable") %>%
                                          dplyr::select(variable, som100, sommeta)) %>%
  dplyr::mutate(sommeta = factor(sommeta)) %>%
  dplyr::mutate(som100 = factor(som100)) %>% 
  dplyr::left_join(marker_genes %>% dplyr::select(gene, population)) %>%
  dplyr::mutate(genepop = paste0(gene, " (", population, ")")) %>%
  dplyr::arrange(population) %>% 
  dplyr::mutate(gene = factor(gene, levels = unique(marker_genes$gene)))

print(ggplot(df, aes(x = sommeta, y = value, fill = sommeta)) + 
        geom_boxplot(outlier.size = 0.5) + facet_wrap(~ genepop, scales = "free_y") + 
        scale_fill_manual(values = cols) + theme_bw())
```

Check the FDRs of the marker genes in each of the clusters

```{r known_markers_fdr, fig.height = 12}
L <- do.call(rbind, lapply(structure(marker_genes$gene, names = marker_genes$gene),
                           function(g) {
  do.call(rbind, lapply(seq_along(scran_markers_all), function(i) {
    j <- grep(paste0("\\.", g, "$"), rownames(scran_markers_all[[i]]))
    if (length(j) == 1) data.frame(gene = g, cluster = i, 
                                   log.FDR = scran_markers_all[[i]][j, "log.FDR"],
                                   stringsAsFactors = FALSE)
    else data.frame(gene = g, cluster = i, log.FDR = NA, stringsAsFactors = FALSE)
  }))
}))
L <- tidyr::spread(L, cluster, log.FDR)
rownames(L) <- L$gene
L$gene <- NULL
L <- L[rowSums(is.na(L)) == 0, ]
pheatmap::pheatmap(exp(L), scale = "none", cluster_rows = TRUE, cluster_cols = TRUE)
```


Plot the individual marker gene expression in each of the 100 original FlowSOM
clusters, color by the final cluster assignment. 

```{r markers-indiv, fig.width = 18}
for (g in unique(df$genepop)) {
  print(ggplot(df %>% dplyr::filter(genepop == g), 
               aes(x = som100, y = value, fill = sommeta)) + 
          geom_boxplot(outlier.size = 0.5) + ggtitle(g) + theme_bw() + 
          theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) + 
          scale_fill_manual(values = cols))
}
```

## Heatmap of marker genes

We make a heatmap of the marker genes, summarized for the cells in each of the 
original FlowSOM clusters. As a summarization function, we use the third
quartile.

```{r, fig.width = 10, fig.height = 12}
dfsum <- df %>% dplyr::group_by(genepop, som100) %>%
  dplyr::summarize(value = quantile(value, probs = 0.75)) %>%
  tidyr::spread(som100, value) %>% as.data.frame()
rownames(dfsum) <- dfsum$genepop
dfsum$genepop <- NULL
dfsum <- dfsum[apply(dfsum, 1, sd) > 0, ]
pheatmap::pheatmap(dfsum, scale = "row", cluster_rows = TRUE, cluster_cols = TRUE, 
                   annotation_col = data.frame(metaClust = factor(metaClustering), row.names = as.character(seq_len(length(metaClustering)))), 
                   show_colnames = FALSE, show_rownames = TRUE, fontsize_row = 6, 
                   annotation_colors = list(metaClust = cols))
```

## t-SNE of subsampled data

Here we apply t-SNE to 10,000 randomly selected cells, using the 50 first PCs as
the input. We color the resulting representation by the assigned cluster. 

```{r tsnesub}
set.seed(123)
subs <- sample(seq_len(nrow(pca)), 10000, replace = FALSE)
pcasub <- pca[subs, ]
rtsne_out <- Rtsne(as.matrix(pcasub), pca = FALSE, verbose = TRUE, perplexity = 30)
rtsne_out <- data.frame(rtsne_out$Y)
colnames(rtsne_out) <- c("tSNE1", "tSNE2")
rtsne_out$sommeta <- factor(colData(sce)$sommeta[subs])
print(ggplot(rtsne_out, aes(x = tSNE1, y = tSNE2, color = sommeta)) + 
        geom_point(size = 0.75) + scale_color_manual(values = cols) + 
        theme_bw())
```

## Session info

```{r}
date()
sessionInfo()
```

