# Normalization for cell-specific biases

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
knitr::opts_chunk$set(dpi=300, dev="png", dev.args=list(pointsize=15))
```

```{r, echo=FALSE, results="hide"}
library(BiocStyle)
library(HDF5Array)
library(scran)
sce <- readRDS("objects/sce.rds")
```

## Performing a rough pre-clustering

Here, we use the deconvolution method to compute size factors for each cell.
We start by doing some pre-clustering to split the cells into sensible clusters, to avoid violating the non-DE assumption and distorting the size factors.

```{r}
library(scran)
system.time({
    ids <- quickCluster(sce, min.mean=0.1, method="igraph",
        block=sce$Library, block.BPPARAM=BPPARAM)
})
summary(tabulate(as.integer(ids)))
```

## Calculating size factors

We now compute a size factor for each cell using the specified parameters.

```{r}
system.time({
    sce <- computeSumFactors(sce, cluster=ids, min.mean=0.1, BPPARAM=BPPARAM)
})
summary(sizeFactors(sce))
```

We can have a look at them in more detail, compared to the library size for each cell.

```{r sizefacplot}
plot(sce$scater_qc$all$total_counts, sizeFactors(sce), 
    log="xy", xlab="Library size", 
    ylab="Size factors", cex=0.2, pch=16)
```

## Normalizing the expression values 

We calculate normalized log-expression values, using delayed operations for speed instead of saving to a new HDF5 file.

```{r}
logcounts(sce) <- log2(t(t(counts(sce))/sizeFactors(sce)) + 1)
```

<!--
```{r echo=FALSE, results="hide"}
# Saving the new object as well.
saveRDS(file="objects/sce.rds", sce)
```
-->
