# Normalization for cell-specific biases

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
knitr::opts_chunk$set(dpi=300, dev="png", dev.args=list(pointsize=15))

library(BiocStyle)
library(HDF5Array)
library(scater)
library(scran)
sce <- readRDS("objects/sce.rds")
```

## Setting up genes and cells

Here, we use the deconvolution method to compute size factors for each cell.
We start by computing the average abundance of each gene.
This is to filter out low-abundance genes, as having too many zeroes will overwhelm the method.
Note that this only applies for normalization - this particular filter is not used in the rest of the analysis.

```{r}
keep <- rowData(sce)$AverageCount >= 0.1
summary(keep)
```

We do some pre-clustering to split the cells into sensible clusters, to avoid violating the non-DE assumption and distorting the size factors.
This is done by applying the `quickCluster` function on the cells in each library.

```{r}
by.lib <- split(seq_len(ncol(sce)), sce$Library)
cluster.id <- character(ncol(sce))
for (lib in names(by.lib)) { 
    current <- by.lib[[lib]]
    ids <- quickCluster(sce[,current], subset.row=keep, method="igraph", 
        max.size=3000, irlba.args=list(maxit=1000))
    cluster.id[current] <- paste0(lib, ".", ids)
}
```

## Calculating size factors

We now compute a size factor for each cell using the specified parameters.

```{r}
library(scran)
sce <- computeSumFactors(sce, cluster=cluster.id, 
    subset.row=keep, min.mean=NULL)
summary(sizeFactors(sce))
```

We can have a look at them in more detail, compared to the library size for each cell.

```{r sizefacplot}
plot(sce$total_counts, sizeFactors(sce), log="xy", xlab="Library size", 
    ylab="Size factors", cex=0.2, pch=16)
```

## Normalizing the expression values 

We calculate normalized log-expression values and save this to a new `HDF5Matrix` object.
This involves resetting the HDF5 dump file directory and location so that the output is retained.

```{r}
setHDF5DumpDir("objects")
setHDF5DumpFile("norm_exprs.h5")
sce <- normalize(sce)
logcounts(sce)
```

We reset the locations to the defaults for future use.

```{r}
setHDF5DumpDir()
setHDF5DumpFile()
```

<!--
```{r echo=FALSE, results="hide"}
# Saving the new object as well.
saveRDS(file="objects/sce.rds", se.out)
```
-->
