# Detecting highly variable genes

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

```{r, echo=FALSE, results="hide"}
library(BiocStyle)
library(HDF5Array)
library(SummarizedExperiment)
se.out <- readRDS("objects/qc_mat.rds")
```

First, we set up the design matrix to block on the library of origin for each cell.
This ensures that we do not detect variance due to technical differences in sequencing.

```{r}
lib.source <- factor(se.out$Library)
design <- model.matrix(~0 + lib.source)
```

Next, we filter out low-abundance genes that do not have enough evidence to reject the null, and just increase the severity of the multiple testing correction.
This also reduces computational work and improves the quality of the fitted trend.

```{r}
library(scater)
ab <- calcAverage(hd5mat, size.factors=se.out$Libsize)
keep <- ab >= 0.01
summary(keep)
```

We compute the variance of the normalized log-expression values while blocking on the library of origin.

```{r}
library(scran)
hd5mat <- assay(se.out, withDimnames=FALSE)
fit <- trendVar(hd5mat, trend="semiloess", design=design, 
                span=0.2, subset.row=keep)
dec <- decomposeVar(fit=fit)
dec <- cbind(rowData(se.out)[keep,1:2], dec)
head(dec)
```

We can examine the mean-variance relationship and verify that the trend was fitted correctly.

```{r}
plot(dec$mean, dec$total, pch=16, cex=0.5, xlab="Mean of log-expression", 
    ylab="Variance of log-expression")
curve(fit$trend(x), add=TRUE, col="red")
```

Saving the results to file.

```{r}
write.table(file="objects/hvg_output.txt", dec, sep="\t", quote=FALSE, col.names=NA)
```
