# Detecting highly variable genes

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

```{r, echo=FALSE, results="hide"}
library(BiocStyle)
library(scran)
sce <- readRDS("objects/sce.rds")
```

First, we set up the design matrix to block on the library of origin for each cell.
This ensures that we do not detect variance due to technical differences in sequencing.

```{r}
lib.source <- factor(sce$Library)
design <- model.matrix(~0 + lib.source)
```

We compute the variance of the normalized log-expression values while blocking on the library of origin, and we fit a trend to it.
Some fiddling with the trend parameters is necessary to obtain an appropriate fit at high abundances.

```{r}
library(scran)
fit <- trendVar(sce, method="loess", parametric=TRUE, 
    design=design, use.spikes=FALSE, 
    loess.args=list(span=0.05, control=loess.control(iterations=100)))
```

We then decompose the biological and technical component for each gene.

```{r}
dec <- decomposeVar(fit=fit)
dec <- cbind(rowData(sce)[,1:2], dec)
dec <- dec[order(dec$p.value, -dec$bio),]
head(dec)
```

We examine the mean-variance relationship and verify that the trend was fitted correctly.

```{r hvgplot}
plot(dec$mean, dec$total, pch=16, cex=0.5, xlab="Mean of log-expression", 
    ylab="Variance of log-expression")
curve(fit$trend(x), add=TRUE, col="red")
```

Finally we save the results to file.

```{r}
write.table(file="objects/hvg_output.txt", dec, sep="\t", quote=FALSE, row.names=FALSE)
```

<!--
Also saving the trend.

```{r}
saveRDS(file="objects/trendfit.rds", fit)
```
-->
