# Detecting highly variable genes

```{r, echo=FALSE, results="hide"}
knitr::opts_chunk$set(error=FALSE, message=FALSE, warning=FALSE)
```

```{r, echo=FALSE, results="hide"}
library(BiocStyle)
library(scran)
sce <- readRDS("objects/sce.rds")

library(DelayedArray)
setAutoBlockSize(100*100*8) # Ensure we load one chunk's worth of rows.
```

We compute the variance of the normalized log-expression values while blocking on the library of origin, and we fit a trend to it.
Some fiddling with the trend parameters is necessary to obtain an appropriate fit at high abundances.

```{r}
system.time({
    fit <- trendVar(sce, method="loess", parametric=TRUE, 
        design=sce$Library, use.spikes=FALSE, BPPARAM=BPPARAM,
        loess.args=list(span=0.05, control=loess.control(iterations=100)))
})
```

For comparison, we fit a trend corresponding to pure Poisson noise.

```{r}
system.time({
    poistrend <- makeTechTrend(
        means=2^seq(0, max(fit$mean), length.out=100) - 1,
        size.factors=sizeFactors(sce),
        approx.npts=10000, BPPARAM=BPPARAM)
})
```

We decompose the biological and technical component for each gene.
Note that we use the Poisson trend here, as the trend fitted to the endogenous variances does not provide a good estimate of the technical noise for UMI data.

```{r}
fit0 <- fit
fit0$trend <- poistrend
dec <- decomposeVar(fit=fit0)
dec <- cbind(rowData(sce)[,1:2], dec)
dec <- dec[order(dec$p.value, -dec$bio),]
head(dec)
```

We examine the mean-variance relationship.

```{r hvgplot}
plot(fit$mean, fit$var, pch=16, cex=0.5, xlab="Mean of log-expression", 
    ylab="Variance of log-expression")
curve(fit$trend(x), add=TRUE, col="red")
curve(poistrend(x), add=TRUE, col="blue")
```

Finally we save the results to file.

```{r}
write.table(file="objects/hvg_output.txt", dec, sep="\t", quote=FALSE, row.names=FALSE)
```

<!--
Also saving the trend.

```{r}
saveRDS(file="objects/trendfit.rds", fit)
```
-->
